### 2024-09-23 paper/paper.qmd
Cleaned data and added docstrings

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

```{r}
data <- read.csv("../data/analysis_data/piccfsa.csv")
```

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-09-23 scripts/01-download_data.R

@@ -1,9 +1,16 @@
+# Purpose: Downloads the data from opendatatoronto API. 
+# Source: https://open.toronto.ca/dataset/persons-in-crisis-calls-for-service-attended/
+# Author: Xinxiang Gao
+# Date: 23 September 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+
 library(opendatatoronto)
 library(dplyr)
 
 # get package
 package <- show_package("persons-in-crisis-calls-for-service-attended")
-package
 
 # get all resources for this package
 resources <- list_package_resources("persons-in-crisis-calls-for-service-attended")
@@ -13,4 +20,5 @@ datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'
 
 # load the first datastore resource as a sample
 data <- filter(datastore_resources, row_number()==1) %>% get_resource()
-write.csv(data, "../data/raw_data/raw_piccfsa.csv", row.names = FALSE)
+write.csv(data, "data/raw_data/raw_piccfsa.csv",  row.names = F)
+

### 2024-09-23 scripts/02-cleaning_data.R

@@ -1,0 +1,15 @@
+# Purpose: Cleans the raw data after download
+# Author: Xinxiang Gao
+# Date: 23 September 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+
+data <- read.csv("data/raw_data/raw_piccfsa.csv")
+
+colnames(data)
+data_1 <- data %>% janitor::clean_names() %>% 
+  select(-x_id, -hood_140, -neighbourhood_140)
+  
+write.csv(data_1, "data/analysis_data/piccfsa.csv", row.names = F)
+  

### 2024-09-24 paper/paper.qmd
Generated some promising plots

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "When and Where: Examining Crisis Service Calls in Toronto"
+subtitle: "A Detailed Look at Temporal and Spatial Trends in Crisis Response Calls"
 author: 
-  - First author
-  - Another author
-thanks: "Code and data are available at: LINK."
+  - Xinxiang Gao
+thanks: "Code and data are available at: https://github.com/xgao28/PICCFSA."
 date: today
@@
 library(tidyverse)
-library(palmerpenguins)
+library(ggplot2)
 ```
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+@citeR, @tidyverse, @ggplot2, @cowplot, @gridExtra
 
-The remainder of this paper is structured as follows. @sec-data....
+@opendatatoronto
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
 ```{r}
@@
 #| echo: false
+#| 
+ggplot(data, aes(x = event_dow, fill = event_type)) + 
+  geom_bar(position = "dodge") + 
+  labs(title = "Event Type Counts Grouped by Day of the Week", 
+       x = "Day of the Week", 
+       y = "Count",
+       fill = "Event Type") +
+  theme_minimal()
+```
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
+```{r}
+ggplot(data, aes(x = event_month, fill = event_type)) + 
+  geom_bar(position = "dodge") + 
+  labs(title = "Event Type Counts Grouped by Day of the Week", 
+       x = "Day of the Week", 
+       y = "Count",
+       fill = "Event Type") +
   theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
   coord_flip()
@@
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
-
-Talk way more about it.
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
+ggplot(data, aes(x = event_year, fill = event_type)) + 
+  geom_bar(position = "dodge") + 
+  labs(title = "Event Type Counts Grouped by Year", 
+       x = "Year", 
+       y = "Count",
+       fill = "Event Type") +
+  theme_minimal()
 ```
@@
 ```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+df_summarized <- data %>%
+  group_by(event_hour, event_type) %>%
+  summarise(count = n()) %>%
+  ungroup()
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
+# Create the line plot
+ggplot(df_summarized, aes(x = event_hour, y = count, color = event_type, group = event_type)) + 
+  geom_line() + 
+  labs(title = "Event Type Counts Grouped by Hour", 
+       x = "Hour", 
+       y = "Count", 
+       color = "Event Type") +
+  theme_minimal()
 ```
 
-# Discussion
+```{r}
+library(cowplot)  # For combining plots
 
-## First discussion point {#sec-first-point}
+# Step 1: Summarize the data by counting occurrences of each event_type
+df_summarized <- data %>%
+  group_by(hood_158, event_type) %>%
+  summarise(count = n()) %>%
+  ungroup()
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+# Step 2: Split the summarized data into separate data frames for each event_type
+event_types <- unique(df_summarized$event_type)
 
-## Second discussion point
+# Create a list to hold the top 10 data frames and plots
+df_list <- list()
+plot_list <- list()
 
-## Third discussion point
+for (event in event_types) {
+  # Filter the top 10 neighborhoods for the current event type
+  df_top10 <- df_summarized %>%
+    filter(event_type == event) %>%
+    arrange(desc(count)) %>%
+    head(10)
 
-## Weaknesses and next steps
+  # Save the data frame to the list
+  df_list[[event]] <- df_top10
 
-Weaknesses and next steps should also be included.
+  # Step 3: Create a bar plot for the top 10 neighborhoods for the current event type
+  plot <- ggplot(df_top10, aes(x = reorder(hood_158, count), y = count)) + 
+    geom_bar(stat = "identity") + 
+    labs(title = event, 
+         x = "Neighborhood 158", 
+         y = "Count") +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+    coord_flip()
 
-\newpage
+  # Save the plot to the list
+  plot_list[[event]] <- plot
+}
 
-\appendix
+# Step 4: Get a look-up table
+neighborhood_table <- data %>%
+  select(hood_158, neighbourhood_158) %>%
+  distinct() %>%
+  arrange(hood_158)
 
-# Appendix {-}
+# Convert the table to a grid object using gridExtra::tableGrob
+table_plot <- gridExtra::tableGrob(neighborhood_table)
 
-# Additional data details
+# Step 5: Combine the individual plots using cowplot
+combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 1)
 
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
+# Step 6: Display the combined plot
+print(combined_plot)
 ```

### 2024-09-24 scripts/01-download_data.R

@@ -16,9 +16,9 @@ package <- show_package("persons-in-crisis-calls-for-service-attended")
 resources <- list_package_resources("persons-in-crisis-calls-for-service-attended")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+datastore_resources <- filter(resources, tolower(format) == 'csv')
 
 # load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+data <- filter(datastore_resources, row_number()==2) %>% get_resource()
 write.csv(data, "data/raw_data/raw_piccfsa.csv",  row.names = F)
 

### 2024-09-24 scripts/02-cleaning_data.R

@@ -7,7 +7,6 @@
 
 data <- read.csv("data/raw_data/raw_piccfsa.csv")
 
-colnames(data)
 data_1 <- data %>% janitor::clean_names() %>% 
   select(-x_id, -hood_140, -neighbourhood_140)
   

### 2024-09-24 paper/paper.qmd
Updated README and plots in paper

@@
 library(ggplot2)
+library(gridExtra)
+library(cowplot)
 ```
@@
 ```{r}
+#| include: false
+#| warning: false
+#| message: false
 data <- read.csv("../data/analysis_data/piccfsa.csv")
@@
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-dow
+#| fig-cap: Event Type Counts Grouped by Day of the Week
 #| echo: false
-#| 
+#| warning: false
+#| message: false
 ggplot(data, aes(x = event_dow, fill = event_type)) + 
@@
 ```{r}
+#| label: fig-month
+#| fig-cap: Event Type Counts Grouped by the Month
+#| echo: false
+#| warning: false
+#| message: false
 ggplot(data, aes(x = event_month, fill = event_type)) + 
   geom_bar(position = "dodge") + 
-  labs(title = "Event Type Counts Grouped by Day of the Week", 
-       x = "Day of the Week", 
+  labs(title = "Event Type Counts Grouped by the Month", 
+       x = "Month", 
        y = "Count",
@@
 
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: fig-year
+#| fig-cap: Event Type Counts Grouped by Year
 #| echo: false
@@
 ```{r}
+#| label: fig-hour
+#| fig-cap: Event Type Counts Grouped by Hour of the Day
+#| echo: false
+#| warning: false
+#| message: false
 df_summarized <- data %>%
@@
 ```{r}
-library(cowplot)  # For combining plots
+#| label: fig-date
+#| fig-cap: Event Type Counts Grouped by Date
+#| echo: false
+#| warning: false
+#| message: false
+df_summarized <- data %>%
+  group_by(event_date, event_type) %>%
+  summarise(count = n()) %>%
+  ungroup()
 
+# Create the line plot
+ggplot(df_summarized, aes(x = event_date, y = count, color = event_type, group = event_type)) + 
+  geom_smooth() + 
+  labs(title = "Event Type Counts Grouped by Date", 
+       x = "Date", 
+       y = "Count", 
+       color = "Event Type") +
+  theme_minimal()
+```
+
+```{r}
+#| label: fig-hood
+#| fig-cap: Event Counts in Neighbourhood Grouped by Event Type
+#| echo: false
+#| warning: false
+#| message: false
 # Step 1: Summarize the data by counting occurrences of each event_type
@@
 plot_list <- list()
+hood_list <- c()
 
@@
   df_list[[event]] <- df_top10
+  hood_list <- c(hood_list, df_top10$hood_158)
 
@@
 
-# Step 4: Get a look-up table
+# Step 4: Combine the individual plots using cowplot
+combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
+
+# Step 5: Display the combined plot
+print(combined_plot)
+```
+
+```{r}
+#| label: tab_hood
+#| fig-cap: A Look-up Table for Hood Number and Hood Name
+#| echo: false
+#| warning: false
+#| message: false
+# Step: Get a look-up table
+hood_set <- unique(hood_list)
 neighborhood_table <- data %>%
@@
   distinct() %>%
-  arrange(hood_158)
+  filter(hood_158 %in% hood_set) %>%
+  arrange(hood_158) %>%
+  rename("hood number" = hood_158,  
+         "hood name" = neighbourhood_158)
 
-# Convert the table to a grid object using gridExtra::tableGrob
-table_plot <- gridExtra::tableGrob(neighborhood_table)
-
-# Step 5: Combine the individual plots using cowplot
-combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 1)
-
-# Step 6: Display the combined plot
-print(combined_plot)
+grid.arrange(grid.table(neighborhood_table))
 ```

### 2024-09-24 scripts/00-simulate_data.R
Updated data simulation

@@ -1,0 +1,37 @@
+# Purpose: Simulates the data similar to the expected dataset. 
+# Author: Xinxiang Gao
+# Date: 23 September 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+# Load necessary libraries
+library(tidyverse)
+library(lubridate)
+
+# Set a seed for reproducibility
+set.seed(690)
+
+# Define parameters for the simulation
+num_rows <- 1000  # Total number of events to simulate
+
+# Generate simulated data
+simulated_data <- data.frame(
+  "_id" = 1:num_rows,
+  OBJECTID = 1:num_rows,
+  EVENT_ID = sample(1000000:27000000, num_rows, replace = FALSE),
+  EVENT_DATE = as.Date("2014-01-01") + sample(0:3000, num_rows, replace = TRUE)
+)
+simulated_data <- simulated_data %>% 
+  mutate(EVENT_YEAR = year(EVENT_DATE),
+         EVENT_MONTH = month(EVENT_DATE),
+         EVENT_DOW = wday(EVENT_DATE, label = TRUE),  # Day of week
+         EVENT_HOUR = sample(0:23, num_rows, replace = TRUE),  # Random hour
+         EVENT_TYPE = sample(c("Suicide-related", "Overdose", "Person in Crisis"), num_rows, replace = TRUE),
+         HOOD_158 = sample(1:174, num_rows, replace = TRUE)) %>% 
+  mutate(NEIGHBOURHOOD_158 = paste("Neighborhood", HOOD_158))
+
+# Print the first few rows of the simulated data to check
+print(head(simulated_data))
+
+# Save the simulated dataset to a CSV file
+write.csv(simulated_data, "data/raw_data/simulated_piccfsa.csv", row.names = FALSE)

### 2024-09-24 paper/paper.qmd
Updated paper and README

@@
 ---
-title: "When and Where: Examining Crisis Service Calls in Toronto"
-subtitle: "A Detailed Look at Temporal and Spatial Trends in Crisis Response Calls"
+title: "When and Where: Understanding the Patterns of Emergency Service Calls in Toronto"
+subtitle: "Exploring the Temporal and Spatial Dynamics of Reported Emergency Incidents"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper examines ten years of data on crisis service calls in Toronto, focusing on overdose, Person in Crisis, and suicide-related incidents.
+Our analysis reveals clear temporal patterns, with overdoses peaking at night and mental health crises spiking in the afternoon.
+Spatially, certain neighborhoods such as Moss Park and Downtown Yonge East consistently report higher numbers of crisis events.
+These findings highlight the importance of targeted mental health and substance abuse interventions, as well as resource allocation strategies tailored to specific times and locations."
 format: pdf
@@
 library(cowplot)
+library(knitr)
 ```
@@
 
-@citeR, @tidyverse, @ggplot2, @cowplot, @gridExtra
+The examination of crisis service calls is critical for enhancing public safety and effectively supporting vulnerable populations in urban environments.
+As mental health issues and substance use continue to rise, it is essential for emergency response services to understand the nature and patterns of crisis situations.
+Analyzing when and where these calls occur can inform strategic planning, resource allocation, and intervention methods for law enforcement and mental health services.
+By shedding light on the dynamics of crisis response, we can better equip first responders to address the complex needs of individuals in distress, ultimately fostering safer and more resilient communities.
 
-@opendatatoronto
+This study aims to analyze the temporal and spatial trends of crisis-related incidents in Toronto, as recorded in the Persons in Crisis Calls for Service dataset. The primary objectives are to:
+
+- Identify the most common types of crisis events and their distribution across neighborhoods.
+- Examine the temporal patterns of these events, including daily, weekly, and seasonal trends.
+
+The sections are organized as follows: 
+@sec-data provides a detailed description of the data and its sources, including the context and ethical considerations.
+@sec-results presents the results using visualizations as well as discusses the implications of the findings, addressing both the practical applications and the limitations of the study.
+Finally, @sec-conclusion concludes the paper and suggests directions for future research.
+
+# Data {#sec-data}
+
+The Persons in Crisis Calls for Service dataset, provided by the Toronto Police Service (TPS), offers a comprehensive view of crisis-related incidents attended by TPS.
+This dataset includes over 309,000 records from January 1, 2014, to June 26, 2024, focusing on significant event types such as Suicide, Person in Crisis, and Overdose.
+The dataset captures incidents that required police intervention, excluding those handled by other TPS divisions, such as Parking and Marine.
+Additionally, the dataset does not contain sensitive information related to individuals in crisis, ensuring that no individual can be identified from the results.
+
+Understanding the nature and patterns of crisis calls is crucial for several reasons.
+For public safety, identifying high-activity areas and time periods can help in deploying resources more effectively, ensuring quicker response times and better support for individuals in crisis.In terms of resource allocation, insights into the demand for crisis services can inform budgeting and staffing decisions, ensuring that resources are allocated where they are most needed.For policy development, the findings can guide the development of policies and interventions aimed at reducing the incidence of crises and improving community well-being.
+
+To provide a comprehensive overview of the data, we use a combination of text, tables, and visualizations.
+The following sections describe the key characteristics of the dataset and summarize the main findings.
+
+# Results {#sec-results}
 
@@
 
-# Data {#sec-data}
-
-```{r}
-#| label: fig-dow
-#| fig-cap: Event Type Counts Grouped by Day of the Week
+```{r summary_stats}
+#| label: tbl-summary
 #| echo: false
@@
 #| message: false
-ggplot(data, aes(x = event_dow, fill = event_type)) + 
-  geom_bar(position = "dodge") + 
-  labs(title = "Event Type Counts Grouped by Day of the Week", 
-       x = "Day of the Week", 
-       y = "Count",
-       fill = "Event Type") +
-  theme_minimal()
+summary_stats <- data %>%
+  group_by(event_type) %>%
+  summarise(
+    total_count = sum(n())
+  ) %>%
+  ungroup()
+
+kable(summary_stats, caption = "Summary Statistics of Crisis Events", align = 'l', booktabs = TRUE)
 ```
 
-```{r}
-#| label: fig-month
-#| fig-cap: Event Type Counts Grouped by the Month
-#| echo: false
-#| warning: false
-#| message: false
-ggplot(data, aes(x = event_month, fill = event_type)) + 
-  geom_bar(position = "dodge") + 
-  labs(title = "Event Type Counts Grouped by the Month", 
-       x = "Month", 
-       y = "Count",
-       fill = "Event Type") +
-  theme_minimal() +
-  coord_flip()
-```
+Over the ten-year period, the dataset reveals significant insights into the nature and frequency of three primary types of crisis events: Overdose, Person in Crisis, and Suicide-related, as summarized in @tbl-summary.
+Among these, Person in Crisis events are by far the most frequent, comprising more than half of the total incidents.
 
-```{r}
-#| label: fig-year
-#| fig-cap: Event Type Counts Grouped by Year
-#| echo: false
-#| warning: false
-#| message: false
+Overdose: There were a total of 34,097 overdose incidents recorded over the decade.
+This category includes all cases where an individual required police intervention due to drug or alcohol overdose.
+The number of overdose incidents highlights the ongoing public health challenge related to substance abuse in the community.
 
-ggplot(data, aes(x = event_year, fill = event_type)) + 
-  geom_bar(position = "dodge") + 
-  labs(title = "Event Type Counts Grouped by Year", 
-       x = "Year", 
-       y = "Count",
-       fill = "Event Type") +
-  theme_minimal()
-```
+Person in Crisis: With 177,851 incidents, Person in Crisis is the most frequently reported type of event.
+These incidents involve individuals who are experiencing mental health crises or other forms of distress that necessitate police involvement.
+The high count underscores the critical need for mental health support and resources within the community.
+
+Suicide-related: A total of 97,155 suicide-related incidents were documented.
+This includes both attempts and situations where there was a risk of suicide, requiring immediate police response.
+The prevalence of such incidents emphasizes the importance of suicide prevention efforts and access to mental health services.
+
+## Temporal Analysis
 
@@
 
+@fig-hour shows the distribution of event types across the hours of the day.
+This plot helps identify any daily patterns in the occurrence of crisis events.
+Overdose incidents, represented in red, peak at night, highlighting the potential need for heightened availability of medical and police services during these hours.
+Person in Crisis events, shown in green, are most frequent in the afternoon and steadily decline toward evening, suggesting that mental health resources may need to be more accessible during business hours.
+Suicide-related events, depicted in blue, follow a trend similar to Overdose events but with lower overall counts, peaking in the late afternoon and early evening hours.
+
+```{r}
+#| label: fig-dow
+#| fig-cap: Event Type Counts Grouped by Day of the Week
+#| echo: false
+#| warning: false
+#| message: false
+data$event_dow <- factor(data$event_dow, 
+    levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+ggplot(data, aes(x = event_dow, fill = event_type)) + 
+  geom_bar(position = "dodge") + 
+  labs(title = "Event Type Counts Grouped by Day of the Week", 
+       x = "Day of the Week", 
+       y = "Count",
+       fill = "Event Type") +
+  theme_minimal()
+```
+
+@fig-dow illustrates the distribution of event types across the days of the week, showing that Person in Crisis events (green bars) consistently dominate the count across all days.
+Thursdays have the highest count of all events, while weekends see a marked drop.
+This could reflect workplace or social stress accumulating during the week, leading to crises toward the week's end.
+The reduction during weekends may point to fewer reports or different response patterns, offering an opportunity for further investigation.
+
+```{r}
+#| label: fig-month
+#| fig-cap: Event Type Counts Grouped by the Month
+#| echo: false
+#| warning: false
+#| message: false
+month_levels <- c("January", "February", "March", "April", "May", "June",
+                  "July", "August", "September", "October", "November", "December")
+
+data$event_month <- factor(data$event_month, levels = month_levels)
+
+ggplot(data, aes(x = event_month, fill = event_type)) + 
+  geom_bar(position = "dodge") + 
+  labs(title = "Event Type Counts Grouped by the Month", 
+       x = "Month", 
+       y = "Count",
+       fill = "Event Type") +
+  theme_minimal() +
+  coord_flip()
+```
+
+@fig-month displays the distribution of event types across the months of the year, revealing seasonal trends.
+Person in Crisis events (green bars) consistently have the highest counts throughout the year, followed by Suicide-related events (blue bars).
+Overdose events (red bars) are the least frequent.
+May and June see the highest number of events across all categories, while February has the lowest, potentially due to fewer outdoor activities during colder months.
+These trends may reflect how seasonal factors and societal conditions influence crisis situations.
+
 ```{r}
@@
 
-# Create the line plot
+df_summarized$event_date <- as.Date(df_summarized$event_date, format = "%Y-%m-%d")
+
 ggplot(df_summarized, aes(x = event_date, y = count, color = event_type, group = event_type)) + 
   geom_smooth() + 
-  labs(title = "Event Type Counts Grouped by Date", 
+  labs(title = "Event Type Counts Over Time", 
        x = "Date", 
@@
        color = "Event Type") +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y",
+               limits = as.Date(c(min(df_summarized$event_date), max(df_summarized$event_date)))) +
   theme_minimal()
 ```
+
+@fig-date presents the temporal trends of crisis calls over the entire period.
+Overall, the chart indicates a general upward trend in the number of crisis events, with a significant peak during the pandemic years of 2021 and 2022, coinciding with lockdowns and restrictions.
+This surge likely reflects the psychological toll of isolation and economic stress, contributing to more frequent mental health crises and overdoses.
+Another noticeable increase is seen in 2024.
+These patterns underscore the importance of mental health and substance abuse support during times of social upheaval.
+
+## Spatial analysis
 
@@
 
-```{r}
-#| label: tab_hood
-#| fig-cap: A Look-up Table for Hood Number and Hood Name
+```{r tab_hood}
+#| label: tbl-hood
 #| echo: false
@@
 
-grid.arrange(grid.table(neighborhood_table))
+kable(neighborhood_table, caption = "A Look-up Table for Hood Number and Hood Name", 
+             booktabs = TRUE, align = 'l', label = "tab-hood")
 ```
+
+@fig-hood showcases the top 10 neighborhoods with the highest counts for each event type.
+Moreover, @tbl-hood provides a lookup table for the neighborhood numbers and their corresponding names, which can be used to interpret the neighborhood codes in the previous plots.
+To illustrate, Moss Park and Downtown Yonge East have the most reported cases for all three event types, with Overdose incidents dominating other neighborhoods.
+The high counts in these areas may be linked to population density, socioeconomic challenges, or proximity to services like shelters and social support networks.
+Kensington-Chinatown ranks third for Person in Crisis events, while Annex ranks third for Suicide-related events.
+These findings suggest that certain neighborhoods may require targeted interventions based on the specific crisis patterns observed.
+
+# Conclusion {#sec-conclusion}
+
+The analysis of the Persons in Crisis Calls for Service dataset reveals meaningful patterns in both the temporal and spatial distribution of crisis-related events in Toronto over the past decade.
+Overdose, Person in Crisis, and Suicide-related incidents exhibit distinct daily, weekly, and seasonal trends.
+Overdose incidents peak at night, Person in Crisis events are most frequent in the afternoon, and Suicide-related incidents show a steady occurrence during the late afternoon and early evening.
+Additionally, Thursdays see the highest volume of crisis calls across all types, while weekends have lower incident counts.
+These temporal patterns provide valuable insight for optimizing resource allocation and tailoring response strategies to times of greatest need.
+
+Spatially, certain neighborhoods, such as Moss Park and Downtown Yonge East, show consistently high rates of crisis incidents, particularly for overdoses.
+These findings highlight the need for targeted interventions in high-risk areas, which could be related to local socioeconomic factors or population density.
+Understanding both the temporal and spatial dynamics of crisis events is crucial for improving emergency response efforts and supporting the communities most affected.
+
+Understanding when and where these crisis events are most likely to occur is critical for resource allocation and emergency response planning.
+By highlighting the neighborhoods and time frames with higher incident rates, this analysis offers actionable insights for policymakers and service providers to improve crisis response strategies, ultimately enhancing public safety and community well-being.
+
+This dataset also contains additional information that could further enrich the analysis, such as the police division of the event, whether an official occurrence was created, and whether a Mental Health Act (MHA) apprehension was made.
+Future work could explore the impact of these factors on crisis management and resolution.
+Additionally, while this dataset focuses on three main crisis types, expanding the scope to include more detailed emergency call categories could offer a more comprehensive understanding of Toronto’s broader crisis response landscape, providing deeper insights into the city's emergency response system.
+
+# Acknowledgements
+
+We would like to express our gratitude to the developers and contributors of R [@citeR] as well as several R packages that were essential for the analysis and visualization of the data in this report.
+The following R packages provided indispensable tools and functionalities:
+
+- tidyverse [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.
+
+- ggplot2 [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.
+
+- cowplot [@cowplot]: The package enabled us to combine multiple plots into a single figure, enhancing the presentation of our results.
+
+- lubridate [@lubridate]: This package made it straightforward to work with date and time data, simplifying the process of manipulating and analyzing temporal information.
+
+- knitr [@knitr]: This package was crucial for integrating R code and its output directly into the document, allowing for reproducible research and dynamic report generation.
+
+Additionally, we are grateful to the City of Toronto and the Open Data Portal for providing access to the dataset through the opendatatoronto package [@opendatatoronto].
+The availability of such high-quality, open-source data is vital for conducting meaningful and impactful research.
+
+Finally, we would like to thank all those who contributed to the development and maintenance of the R programming language and its ecosystem, as well as the broader open-source community, whose efforts make such research possible.
 

### 2024-09-26 paper/paper.qmd
Changed the code structure
Added test-data.

Moved some data cleaning from paper to cleaning-data.R.

Minor updates: Updated reference, simulate_data

@@
 #| message: false
-data$event_dow <- factor(data$event_dow, 
-    levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
 
@@
 #| message: false
-month_levels <- c("January", "February", "March", "April", "May", "June",
-                  "July", "August", "September", "October", "November", "December")
-
-data$event_month <- factor(data$event_month, levels = month_levels)
-
 ggplot(data, aes(x = event_month, fill = event_type)) + 
@@
 
+- janitor [@janitor]: This package provided helpful tools for data cleaning. Its functionalities greatly enhanced the efficiency and accuracy of our data preparation process.
+
 Additionally, we are grateful to the City of Toronto and the Open Data Portal for providing access to the dataset through the opendatatoronto package [@opendatatoronto].

### 2024-09-26 scripts/00-simulate_data.R

@@ -16,15 +16,14 @@ num_rows <- 1000  # Total number of events to simulate
 
 # Generate simulated data
 simulated_data <- data.frame(
-  "_id" = 1:num_rows,
   OBJECTID = 1:num_rows,
   EVENT_ID = sample(1000000:27000000, num_rows, replace = FALSE),
   EVENT_DATE = as.Date("2014-01-01") + sample(0:3000, num_rows, replace = TRUE)
 )
 simulated_data <- simulated_data %>% 
   mutate(EVENT_YEAR = year(EVENT_DATE),
-         EVENT_MONTH = month(EVENT_DATE),
-         EVENT_DOW = wday(EVENT_DATE, label = TRUE),  # Day of week
+         EVENT_MONTH = month(EVENT_DATE, label = TRUE, abbr = FALSE),
+         EVENT_DOW = wday(EVENT_DATE, label = TRUE, abbr = FALSE),  # Day of week
          EVENT_HOUR = sample(0:23, num_rows, replace = TRUE),  # Random hour
          EVENT_TYPE = sample(c("Suicide-related", "Overdose", "Person in Crisis"), num_rows, replace = TRUE),
          HOOD_158 = sample(1:174, num_rows, replace = TRUE)) %>% 

### 2024-09-26 scripts/01-download_data.R

@@ -15,10 +15,10 @@ package <- show_package("persons-in-crisis-calls-for-service-attended")
 # get all resources for this package
 resources <- list_package_resources("persons-in-crisis-calls-for-service-attended")
 
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+# identify datastore resources
 datastore_resources <- filter(resources, tolower(format) == 'csv')
 
-# load the first datastore resource as a sample
+# load the second datastore resource
 data <- filter(datastore_resources, row_number()==2) %>% get_resource()
 write.csv(data, "data/raw_data/raw_piccfsa.csv",  row.names = F)
 

### 2024-09-26 scripts/02-cleaning_data.R

@@ -4,11 +4,22 @@
 # Contact: xinxiang.gao@mail.utoronto.ca
 # License: MIT
 
-
+# read in the raw data
 data <- read.csv("data/raw_data/raw_piccfsa.csv")
 
+# removed unnecessary columns
 data_1 <- data %>% janitor::clean_names() %>% 
   select(-x_id, -hood_140, -neighbourhood_140)
+
+# Factor day of week and month for correct ordering
+data_1$event_dow <- factor(data_1$event_dow, 
+                         levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+month_levels <- c("January", "February", "March", "April", "May", "June",
+                  "July", "August", "September", "October", "November", "December")
+
+data_1$event_month <- factor(data_1$event_month, levels = month_levels)
   
+# outoput as processed data
 write.csv(data_1, "data/analysis_data/piccfsa.csv", row.names = F)
   

### 2024-09-26 scripts/03-test_data.R

@@ -1,0 +1,31 @@
+# Purpose: Test and validate the simulated data 
+# Author: Xinxiang Gao
+# Date: 26 September 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+library(tidyverse)
+library(lubridate)
+
+data <- read.csv("data/raw_data/simulated_piccfsa.csv")
+
+# date in range
+start_time <- as_date("2014-01-01")
+end_time <- as_date("2024-06-30")
+test_date_range <- all(data$EVENT_DATE >= start_time & data$EVENT_DATE <= end_time)
+
+# months are correctly generated
+month_levels <- c("January", "February", "March", "April", "May", "June",
+                  "July", "August", "September", "October", "November", "December")
+test_month_range <- all(data$EVENT_MONTH %in% month_levels)
+
+# day of weeks are correctly generated
+dow_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
+test_dow_range <- all(data$EVENT_DOW %in% dow_levels)
+
+# types are correctly generated
+type_levels <- c("Suicide-related", "Overdose", "Person in Crisis")
+test_type_range <- all(data$EVENT_TYPE %in% type_levels)
+
+print(c(test_date_range, test_month_range, test_dow_range, test_type_range))
+

### 2024-09-27 paper/paper.qmd
Added removing NA observations
and updated accordingly

@@
 date-format: long
+toc: true
 abstract: "This paper examines ten years of data on crisis service calls in Toronto, focusing on overdose, Person in Crisis, and suicide-related incidents.
@@
 The Persons in Crisis Calls for Service dataset, provided by the Toronto Police Service (TPS), offers a comprehensive view of crisis-related incidents attended by TPS.
-This dataset includes over 309,000 records from January 1, 2014, to June 26, 2024, focusing on significant event types such as Suicide, Person in Crisis, and Overdose.
+This dataset includes over 309,000 records from January 1, 2014, to June 30, 2024, focusing on significant event types such as Suicide, Person in Crisis, and Overdose.
 The dataset captures incidents that required police intervention, excluding those handled by other TPS divisions, such as Parking and Marine.
@@
 
-Understanding the nature and patterns of crisis calls is crucial for several reasons.
-For public safety, identifying high-activity areas and time periods can help in deploying resources more effectively, ensuring quicker response times and better support for individuals in crisis.In terms of resource allocation, insights into the demand for crisis services can inform budgeting and staffing decisions, ensuring that resources are allocated where they are most needed.For policy development, the findings can guide the development of policies and interventions aimed at reducing the incidence of crises and improving community well-being.
+Understanding the nature and patterns of crisis calls is foundational for several reasons.
+For public safety, identifying high-activity areas and time periods can help in deploying resources more effectively, ensuring quicker response times and better support for individuals in crisis.In terms of resource allocation, statistical results towards the demand for crisis services can inform budgeting and staffing decisions, ensuring that resources are allocated where they are most needed.For policy development, the findings can guide the development of policies and interventions aimed at reducing the incidence of crises and improving community well-being.
 
@@
 data <- read.csv("../data/analysis_data/piccfsa.csv")
+
+# Factor day of week and month for correct ordering
+data$event_dow <- factor(data$event_dow, 
+                         levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+month_levels <- c("January", "February", "March", "April", "May", "June",
+                  "July", "August", "September", "October", "November", "December")
+
+data$event_month <- factor(data$event_month, levels = month_levels)
 ```
@@
   ) %>%
-  ungroup()
+  ungroup() %>% 
+  rename("Event Type" = event_type, "Total Count" = total_count)
 
@@
 
-Over the ten-year period, the dataset reveals significant insights into the nature and frequency of three primary types of crisis events: Overdose, Person in Crisis, and Suicide-related, as summarized in @tbl-summary.
+## Overview
+Over the ten-year period, the dataset reveals the nature and frequency of three primary types of crisis events: Overdose, Person in Crisis, and Suicide-related, as summarized in @tbl-summary.
 Among these, Person in Crisis events are by far the most frequent, comprising more than half of the total incidents.
 
-Overdose: There were a total of 34,097 overdose incidents recorded over the decade.
+- Overdose: There were a total of 32,494 overdose incidents recorded over the decade.
 This category includes all cases where an individual required police intervention due to drug or alcohol overdose.
@@
 
-Person in Crisis: With 177,851 incidents, Person in Crisis is the most frequently reported type of event.
+- Person in Crisis: With 172,149 incidents, Person in Crisis is the most frequently reported type of event.
 These incidents involve individuals who are experiencing mental health crises or other forms of distress that necessitate police involvement.
@@
 
-Suicide-related: A total of 97,155 suicide-related incidents were documented.
+- Suicide-related: A total of 92,762 suicide-related incidents were documented.
 This includes both attempts and situations where there was a risk of suicide, requiring immediate police response.
@@
 #| label: fig-hour
-#| fig-cap: Event Type Counts Grouped by Hour of the Day
+#| fig-cap: Event Type Counts by Hour of the Day
 #| echo: false
@@
   geom_line() + 
-  labs(title = "Event Type Counts Grouped by Hour", 
+  labs(title = "Event Type Counts by Hour", 
        x = "Hour", 
@@
 #| label: fig-dow
-#| fig-cap: Event Type Counts Grouped by Day of the Week
+#| fig-cap: Event Type Counts by Day of the Week
 #| echo: false
@@
   geom_bar(position = "dodge") + 
-  labs(title = "Event Type Counts Grouped by Day of the Week", 
+  labs(title = "Event Type Counts by Day of the Week", 
        x = "Day of the Week", 
@@
        fill = "Event Type") +
-  theme_minimal()
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
 ```
@@
 #| label: fig-month
-#| fig-cap: Event Type Counts Grouped by the Month
+#| fig-cap: Event Type Counts by the Month
 #| echo: false
@@
   geom_bar(position = "dodge") + 
-  labs(title = "Event Type Counts Grouped by the Month", 
+  labs(title = "Event Type Counts by the Month", 
        x = "Month", 
@@
 #| label: fig-date
-#| fig-cap: Event Type Counts Grouped by Date
+#| fig-cap: Event Type Counts Over Time
 #| echo: false
@@
                limits = as.Date(c(min(df_summarized$event_date), max(df_summarized$event_date)))) +
-  theme_minimal()
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
@@
 #| label: fig-hood
-#| fig-cap: Event Counts in Neighbourhood Grouped by Event Type
+#| fig-cap: Event Counts in Neighbourhood by Event Type
 #| echo: false
@@
 
+# Define colors for each event type
+event_colors <- c("Overdose" = "#1B9E77",      # Deep green
+                  "Person in Crisis" = "#D95F02",  # Muted orange
+                  "Suicide-related" = "#7570B3")   # Slate blue
+
 for (event in event_types) {
@@
 
-  # Step 3: Create a bar plot for the top 10 neighborhoods for the current event type
+  # Step 3: Create a bar plot with a consistent color for each event type
   plot <- ggplot(df_top10, aes(x = reorder(hood_158, count), y = count)) + 
-    geom_bar(stat = "identity") + 
+    geom_bar(stat = "identity", fill = event_colors[event]) +  # Set a fixed color for the bars
     labs(title = event, 
@@
 
-kable(neighborhood_table, caption = "A Look-up Table for Hood Number and Hood Name", 
+kable(neighborhood_table, caption = "Hood Number and Name Reference", 
              booktabs = TRUE, align = 'l', label = "tab-hood")
@@
 
-The analysis of the Persons in Crisis Calls for Service dataset reveals meaningful patterns in both the temporal and spatial distribution of crisis-related events in Toronto over the past decade.
-Overdose, Person in Crisis, and Suicide-related incidents exhibit distinct daily, weekly, and seasonal trends.
-Overdose incidents peak at night, Person in Crisis events are most frequent in the afternoon, and Suicide-related incidents show a steady occurrence during the late afternoon and early evening.
-Additionally, Thursdays see the highest volume of crisis calls across all types, while weekends have lower incident counts.
-These temporal patterns provide valuable insight for optimizing resource allocation and tailoring response strategies to times of greatest need.
+The analysis of the Persons in Crisis Calls for Service dataset shows notable patterns in the timing and location of crisis-related incidents in Toronto over the past decade.
+Overdose, Person in Crisis, and Suicide-related events each follow distinct daily, weekly, and seasonal cycles.
+Overdoses tend to occur most frequently at night, Person in Crisis calls are most common in the afternoon, and Suicide-related incidents maintain a steady rate during late afternoon and early evening.
+Thursdays experience the highest volume of crisis calls across all types, while weekends typically have fewer incidents.
+These time-based trends help guide the allocation of resources and the adjustment of response plans to align with periods of greatest demand.
 
-Spatially, certain neighborhoods, such as Moss Park and Downtown Yonge East, show consistently high rates of crisis incidents, particularly for overdoses.
-These findings highlight the need for targeted interventions in high-risk areas, which could be related to local socioeconomic factors or population density.
-Understanding both the temporal and spatial dynamics of crisis events is crucial for improving emergency response efforts and supporting the communities most affected.
+In terms of location, areas like Moss Park and Downtown Yonge East consistently report high numbers of crisis calls, particularly for overdoses.
+These findings suggest the need for focused interventions in these neighborhoods, potentially influenced by socioeconomic conditions or population density.
+Recognizing both the temporal and geographic patterns of crisis incidents is key to enhancing emergency response strategies and supporting the areas most impacted.
 
-Understanding when and where these crisis events are most likely to occur is critical for resource allocation and emergency response planning.
-By highlighting the neighborhoods and time frames with higher incident rates, this analysis offers actionable insights for policymakers and service providers to improve crisis response strategies, ultimately enhancing public safety and community well-being.
+Knowing when and where these incidents are most likely to happen is vital for planning emergency response efforts and distributing resources effectively.
+By identifying neighborhoods and time periods with higher incident rates, this analysis offers practical guidance for policymakers and service providers to improve crisis management, ultimately contributing to public safety and community well-being.
 
-This dataset also contains additional information that could further enrich the analysis, such as the police division of the event, whether an official occurrence was created, and whether a Mental Health Act (MHA) apprehension was made.
-Future work could explore the impact of these factors on crisis management and resolution.
-Additionally, while this dataset focuses on three main crisis types, expanding the scope to include more detailed emergency call categories could offer a more comprehensive understanding of Toronto’s broader crisis response landscape, providing deeper insights into the city's emergency response system.
+The dataset also includes additional information that could deepen the analysis, such as police division details, whether an official occurrence was logged, and whether a Mental Health Act (MHA) apprehension occurred.
+Future studies could examine how these elements influence crisis resolution.
+Moreover, although this dataset focuses on three crisis types, incorporating a broader range of emergency call categories could provide a more detailed understanding of Toronto's crisis response landscape.
 
@@
 
-- tidyverse [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.
+- `tidyverse` [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.
 
-- ggplot2 [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.
+- `ggplot2` [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.
 
-- cowplot [@cowplot]: The package enabled us to combine multiple plots into a single figure, enhancing the presentation of our results.
+- `cowplot` [@cowplot]: The package enabled us to combine multiple plots into a single figure, enhancing the presentation of our results.
 
-- lubridate [@lubridate]: This package made it straightforward to work with date and time data, simplifying the process of manipulating and analyzing temporal information.
+- `lubridate` [@lubridate]: This package facilitates the work with date and time data, simplifying the process of manipulating and analyzing temporal information.
 
-- knitr [@knitr]: This package was crucial for integrating R code and its output directly into the document, allowing for reproducible research and dynamic report generation.
+- `knitr` [@knitr]: This package enabled us to perform data demonstration with tables.
 
-- janitor [@janitor]: This package provided helpful tools for data cleaning. Its functionalities greatly enhanced the efficiency and accuracy of our data preparation process.
+- `janitor` [@janitor]: This package provided helpful tools for data cleaning. Its functionalities greatly enhanced the efficiency and accuracy of our data preparation process.
 
@@
 
+# Appendix
+
+## Data Cleaning
+
+In the data cleaning process, the raw dataset was first read from a CSV file and unnecessary columns, including identifiers and irrelevant variables, were removed.
+The janitor package was used to clean the column names for consistency.
+Instances of the value "NSA" in character columns were replaced with NA, and any rows containing NA values were subsequently dropped.
+To ensure proper ordering in visualizations, the day of the week and month were converted to factors with defined levels.
+Finally, the cleaned dataset was saved as a new CSV file for further analysis.
+The cleaned dataset contains over 297,000 observations of 10 variables.
+
 \newpage

### 2024-09-27 scripts/00-simulate_data.R

@@ -11,26 +11,25 @@ library(lubridate)
 # Set a seed for reproducibility
 set.seed(690)
 
-# Define parameters for the simulation
 num_rows <- 1000  # Total number of events to simulate
 
 # Generate simulated data
 simulated_data <- data.frame(
-  OBJECTID = 1:num_rows,
-  EVENT_ID = sample(1000000:27000000, num_rows, replace = FALSE),
-  EVENT_DATE = as.Date("2014-01-01") + sample(0:3000, num_rows, replace = TRUE)
-)
-simulated_data <- simulated_data %>% 
-  mutate(EVENT_YEAR = year(EVENT_DATE),
-         EVENT_MONTH = month(EVENT_DATE, label = TRUE, abbr = FALSE),
-         EVENT_DOW = wday(EVENT_DATE, label = TRUE, abbr = FALSE),  # Day of week
-         EVENT_HOUR = sample(0:23, num_rows, replace = TRUE),  # Random hour
-         EVENT_TYPE = sample(c("Suicide-related", "Overdose", "Person in Crisis"), num_rows, replace = TRUE),
-         HOOD_158 = sample(1:174, num_rows, replace = TRUE)) %>% 
-  mutate(NEIGHBOURHOOD_158 = paste("Neighborhood", HOOD_158))
+  objectid = 1:num_rows,
+  event_id = sample(1000000:27000000, num_rows, replace = FALSE),
+  event_date = as.Date("2014-01-01") + sample(0:3000, num_rows, replace = TRUE)
+  ) %>% 
+  mutate(event_year = year(event_date),
+         event_month = month(event_date, label = TRUE, abbr = FALSE),
+         event_dow = wday(event_date, label = TRUE, abbr = FALSE),  # Day of week
+         event_hour = sample(0:23, num_rows, replace = TRUE),  
+         event_type = sample(c("Suicide-related", "Overdose", "Person in Crisis"), num_rows, replace = TRUE),
+         hood_158 = sample(1:174, num_rows, replace = TRUE)) %>% 
+  mutate(neighbourhood_158 = paste("Neighborhood", hood_158))
 
 # Print the first few rows of the simulated data to check
 print(head(simulated_data))
 
 # Save the simulated dataset to a CSV file
 write.csv(simulated_data, "data/raw_data/simulated_piccfsa.csv", row.names = FALSE)
+

### 2024-09-27 scripts/02-cleaning_data.R

@@ -7,9 +7,11 @@
 # read in the raw data
 data <- read.csv("data/raw_data/raw_piccfsa.csv")
 
-# removed unnecessary columns
+# removed unnecessary columns and NA observations
 data_1 <- data %>% janitor::clean_names() %>% 
-  select(-x_id, -hood_140, -neighbourhood_140)
+  select(-x_id, -hood_140, -neighbourhood_140, -division, -occurrence_created, -apprehension_made) %>% 
+  mutate(across(where(is.character), ~ na_if(., "NSA"))) %>% 
+  drop_na()
 
 # Factor day of week and month for correct ordering
 data_1$event_dow <- factor(data_1$event_dow, 

### 2024-09-27 scripts/03-test_data.R

@@ -7,6 +7,8 @@
 library(tidyverse)
 library(lubridate)
 
+## simulated data test
+
 data <- read.csv("data/raw_data/simulated_piccfsa.csv")
 
 # date in range
@@ -26,6 +28,30 @@ test_dow_range <- all(data$EVENT_DOW %in% dow_levels)
 # types are correctly generated
 type_levels <- c("Suicide-related", "Overdose", "Person in Crisis")
 test_type_range <- all(data$EVENT_TYPE %in% type_levels)
+
+print(c(test_date_range, test_month_range, test_dow_range, test_type_range))
+
+
+# cleaned data test
+data <- read.csv("data/analysis_data/piccfsa.csv")
+
+# date in range
+start_time <- as_date("2014-01-01")
+end_time <- as_date("2024-06-30")
+test_date_range <- all(data$event_date >= start_time & data$event_date <= end_time)
+
+# months are correctly generated
+month_levels <- c("January", "February", "March", "April", "May", "June",
+                  "July", "August", "September", "October", "November", "December")
+test_month_range <- all(data$event_month %in% month_levels)
+
+# day of weeks are correctly generated
+dow_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
+test_dow_range <- all(data$event_dow %in% dow_levels)
+
+# types are correctly generated
+type_levels <- c("Suicide-related", "Overdose", "Person in Crisis")
+test_type_range <- all(data$event_type %in% type_levels)
 
 print(c(test_date_range, test_month_range, test_dow_range, test_type_range))
 

### 2024-09-27 paper/paper.qmd
Changed the paper structure
Moved acknowledgements to appendix.
Added literature to sociologically discuss the approaches.

@@
 
-The sections are organized as follows: 
+The main sections are organized as follows: 
 @sec-data provides a detailed description of the data and its sources, including the context and ethical considerations.
 @sec-results presents the results using visualizations as well as discusses the implications of the findings, addressing both the practical applications and the limitations of the study.
-Finally, @sec-conclusion concludes the paper and suggests directions for future research.
+@sec-conclusion concludes the paper and suggests directions for future research.
 
@@
 
-## Overview
+## Result Overview
 Over the ten-year period, the dataset reveals the nature and frequency of three primary types of crisis events: Overdose, Person in Crisis, and Suicide-related, as summarized in @tbl-summary.
@@
 These findings suggest the need for focused interventions in these neighborhoods, potentially influenced by socioeconomic conditions or population density.
-Recognizing both the temporal and geographic patterns of crisis incidents is key to enhancing emergency response strategies and supporting the areas most impacted.
+Notably, Moss Park is one of Toronto’s main encampment sites, and the research on pandemic encampments by @edhi2022pandemic provides a complementary perspective on the broader social context within which these crises occur.
+The study highlights that the City of Toronto has been experiencing one of its deepest housing crises, with homeless encampments serving as sites of resistance against systemic violence and insufficient interventions.
+The displacement and criminalization of unhoused residents during the pandemic underscore the intersectionality of housing insecurity and mental health crises.
+As such, addressing the root causes of homelessness, including the lack of affordable housing and supportive services, could contribute to reducing the frequency of crisis calls, especially in areas with high concentrations of vulnerable populations.
+Integrating a human rights-based approach to housing and crisis management, as advocated by Edhi, can provide a more holistic framework for responding to and preventing these emergencies, thereby fostering a more equitable and resilient urban environment.
 
@@
 
-The dataset also includes additional information that could deepen the analysis, such as police division details, whether an official occurrence was logged, and whether a Mental Health Act (MHA) apprehension occurred.
+In terms of future work, the dataset also includes additional information that could deepen the analysis, such as police division details, whether an official occurrence was logged, and whether a Mental Health Act (MHA) apprehension occurred.
 Future studies could examine how these elements influence crisis resolution.
@@
 
-# Acknowledgements
+# Appendix {#sec-app}
+
+## Acknowledgements
 
@@
 
-# Appendix
-
 ## Data Cleaning

### 2024-09-27 paper/paper.qmd
Updated paper
Added `styler`.
Added footnote to data cleaning section.
Added motivation and variable definition to data cleaning section.

@@
 library(knitr)
+library(styler)
 ```
@@
 Analyzing when and where these calls occur can inform strategic planning, resource allocation, and intervention methods for law enforcement and mental health services.
-By shedding light on the dynamics of crisis response, we can better equip first responders to address the complex needs of individuals in distress, ultimately fostering safer and more resilient communities.
+By discovering on the dynamics of crisis response, we can better equip first responders to address the complex needs of individuals in distress, ultimately fostering safer and more resilient communities.
 
@@
 - Identify the most common types of crisis events and their distribution across neighborhoods.
-- Examine the temporal patterns of these events, including daily, weekly, and seasonal trends.
+- Examine the temporal patterns of these events, including daily, weekly, seasonal and long-term trends.
 
 The main sections are organized as follows: 
-@sec-data provides a detailed description of the data and its sources, including the context and ethical considerations.
-@sec-results presents the results using visualizations as well as discusses the implications of the findings, addressing both the practical applications and the limitations of the study.
-@sec-conclusion concludes the paper and suggests directions for future research.
+@sec-data provides a detailed description of the data and its sources, including the context, limitations and ethical considerations.
+@sec-results presents the analysis through a series of figures and tables, interpreting key patterns in the timing, location, and frequency of crisis-related incidents, and discussing their practical implications.
+@sec-conclusion summarizes the findings, connects them with relevant literature on housing insecurity and mental health, and discusses how this analysis can inform future research and policy decisions for crisis management.
 
@@
 
-The Persons in Crisis Calls for Service dataset, provided by the Toronto Police Service (TPS), offers a comprehensive view of crisis-related incidents attended by TPS.
+## Data Overview
+
+The Persons in Crisis Calls for Service dataset offers a comprehensive view of crisis-related incidents attended by TPS.
 This dataset includes over 309,000 records from January 1, 2014, to June 30, 2024, focusing on significant event types such as Suicide, Person in Crisis, and Overdose.
@@
 
+The dataset consists of 10 primary variables after cleaning[^1], focusing on unique identifiers and categorizing event types, along with associated spatial and temporal information.
+
+[^1]: A comprehensive description of the data cleaning process, including the procedures, motivations, and variable definitions, can be found in [Section 5.2](#sec-dc).
+
+## Data Collection Process
+Each entry in the dataset reflects a real-world crisis event attended by TPS officers.
+The data are based on police dispatch logs, meaning that a record is generated whenever officers respond to an event.
+The event types represent a wide range of crises, such as suicide attempts or overdose situations, and the TPS categorizes each event based on agency-specified fields.
+
+The dataset is queried by the date the event was reported, not necessarily the date of occurrence, which may introduce a slight reporting delay for some incidents.
+Privacy considerations, particularly the offsetting of locations to the nearest road intersection node, limit geographic precision.
+While the dataset does include neighbourhood and police division information, these are approximate.
+
+## Practical Implications
 Understanding the nature and patterns of crisis calls is foundational for several reasons.
 For public safety, identifying high-activity areas and time periods can help in deploying resources more effectively, ensuring quicker response times and better support for individuals in crisis.In terms of resource allocation, statistical results towards the demand for crisis services can inform budgeting and staffing decisions, ensuring that resources are allocated where they are most needed.For policy development, the findings can guide the development of policies and interventions aimed at reducing the incidence of crises and improving community well-being.
-
-To provide a comprehensive overview of the data, we use a combination of text, tables, and visualizations.
-The following sections describe the key characteristics of the dataset and summarize the main findings.
 
@@
 
+- `styler` [@styler]: This package is helpful for styling the code.
+
 Additionally, we are grateful to the City of Toronto and the Open Data Portal for providing access to the dataset through the opendatatoronto package [@opendatatoronto].
@@
 
-## Data Cleaning
+## Data Cleaning {#sec-dc}
 
-In the data cleaning process, the raw dataset was first read from a CSV file and unnecessary columns, including identifiers and irrelevant variables, were removed.
+In the data cleaning process, the raw dataset was first read from a CSV file and irrelevant variables, were removed.
 The janitor package was used to clean the column names for consistency.
@@
 Finally, the cleaned dataset was saved as a new CSV file for further analysis.
-The cleaned dataset contains over 297,000 observations of 10 variables.
+The cleaned dataset contains over 297,000 observations of 10 variables:
+
+- `objectid`: A unique identifier for each record.
+- `event_id`: A numeric identifier for the event, distinct from the object ID.
+- `event_date`: The exact date when the event occurred, used for time-based analysis.
+- `event_year`: The year in which the event occurred, facilitating yearly trend analysis.
+- `event_month`: The month of the event, converted into an ordered factor for proper ordering in visualizations.
+- `event_dow`: The day of the week on which the event occurred, also converted to an ordered factor.
+- `event_hour`: The hour the call was received, for discovering time-of-day patterns for crisis incidents.
+- `event_type`: A field that categorizes the event, with possible values including Suicide-related, Person in Crisis, and Overdose.
+- `hood_158`: The identifier of the neighbourhood using the City of Toronto’s new 158-neighbourhood structure.
+- `neighbourhood_158`: The name of the neighbourhood, corresponding to the new 158-neighbourhood structure.
+
+In terms of unnecessary columns, the raw dataset contains several additional variables, including `DIVISION`, `OCCURRENCE_CREATED`, `APPREHENSION_MADE`, `HOOD_140`, and `NEIGHBOURHOOD_140`, which were not the focus of our research.
+Our analysis centers specifically on the temporal and spatial patterns of crisis incidents rather than the effects of intervention by TPS.
+Consequently, we chose to remove these columns to streamline our dataset and concentrate on the variables most relevant to our research objectives.
+Specifically, we opted to exclude `HOOD_140` and `NEIGHBOURHOOD_140` in favor of the newer 158-neighbourhood labeling system to ensure our analysis reflects the most current geographic designations.
 
